---
layout: default
title: Tutorial
published: true
submenu: tutorial
---

Binary, ordinal and nominal variables are considered categorical (not
continuous). It makes a big difference if these categorical variables are
exogenous (independent) or endogenous (dependent) in the model.

#### Exogenous categorical variables ####

If you have a binary exogenous covariate (say, gender), all you need to do is
to recode it as a dummy (0/1) variable. Just like you would do in a classic
regression model. If you have an exogenous ordinal variable, you can use a
coding scheme reflecting the order (say, 1,2,3,...) and treat it as any other
(numeric) covariate. If you have a nominal categorical variable with $K > 2$
levels, you need to replace it by a set of $K-1$ dummy variables, again, just
like you would do in classical regression.

#### Endogenous categorical variables ####

The lavaan 0.5 series can deal with binary and ordinal (but not nominal)
endogenous variables. There are two ways to communicate to lavaan that
some of the endogenous variables are to be treated as categorical:

1. declare them as 'ordered' (using the `ordered` function, which is part of
base R) in your data.frame before you run the analysis; for example, if you
need to declare four variables (say, `item1`, `item2`, `item3`, `item4`) as
ordinal in your data.frame (called `Data`), you can use something like:

    ```{r eval=FALSE, tidy=FALSE}
    Data[,c("item1",
            "item2",
            "item3",
            "item4")] <-
        lapply(Data[,c("item1",
                       "item2",
                       "item3",
                       "item4")], ordered)
    ```
2. use the `ordered` argument when using one of the fitting functions
(cfa/sem/growth/lavaan), for example, if you have four binary or ordinal
variables (say, `item1`, `item2`, `item3`, `item4`), you can use:

    ```{r eval=FALSE, tidy=FALSE}
    fit <- cfa(myModel, data = myData,
               ordered = c("item1","item2",
                           "item3","item4"))
    ```

If all the (endogenous) variables are to be treated as categorical, you can
use `ordered = TRUE` as a shortcut.

When the `ordered=` argument is used, lavaan will automatically switch to the
`WLSMV` estimator: it will use diagonally weighted least squares (`DWLS`) to
estimate the model parameters, but it will use the full weight matrix to
compute robust standard errors, and a mean- and variance-adjusted test
stastistic. Other options are unweighted least squares (ULSMV), or pairwise
maximum likelihood (PML). Full information maximum likelihood is currently not
supported.


