---
layout: default
title: Tutorial
published: true
submenu: tutorial
---

If a measurement model contains multiple latent variables (factors), we usually
know which indicators belong to each factor. We call this the factor structure.
Confirmatory factor analysis can be used to check if this a priori factor
structure holds in the data. There are settings, however, where the factor
structure is unclear, and we wish to rotate the solution in order to find a
suitable structure in a given model. When the model also includes a structural
part (i.e., regressions among the latent variables), this is referred to as
exploratory structural equation modeling or ESEM. If there is only a
measurement part, this is called exploratory factor analysis (EFA). What they
have in common is that the factor structure (for one or more blocks) is found
by means of rotation.

```{r echo=FALSE, message=FALSE}
library(lavaan)
```


#### ESEM ####

To illustrate how ESEM works in lavaan, consider the following syntax:

```{r eval=TRUE, tidy=FALSE}
model <- '
    # efa block 1
    efa("efa1")*f1 + 
    efa("efa1")*f2 =~ x1 + x2 + x3 + x4 + x5 + x6

    # efa block 2
    efa("efa2")*f3 + 
    efa("efa2")*f4 =~ y1 + y2 + y3 + y4 + y5 + y6

    # cfa block
    f5 =~ z7 + z8 + z9
    f6 =~ z10 + z11 + z12

    # regressions
    f3 ~ f1 + f2
    f4 ~ f3
'
```
This model syntax defines six latent variables (or factors).  For f5 and f6,
the factor structure is known, and they belong to a regular CFA block. But for
f1 and f2, the factor structure is not known, and we will use a rotation method
to find an appropiate structure. The f1 and f2 factors belong together in an
EFA block that is (arbitrarily) named `efa1`. The `efa("efa1")*` modifier just
before f1 and f2 is used to alert lavaan that these two factors belong to the
same EFA block. The factors f3 and f4 belong to a different EFA block
(named `efa2`) and will be rotated independently. The structural part of the
model is given as usual. To fit this model, we could call the `sem()` function
as follows:


```{r eval=FALSE, tidy=FALSE}
fit <- sem(model = model, data = myData, rotation = "geomin")
```

Different rotation criteria are available, and many rotation options can 
be provided (see the manual page for the `efa()` function for an overview).

To illustrate ESEM, we will borrow an example from the Mplus User's Guide
(example 5.25). First we read in the data:

```{r eval=TRUE, tidy=FALSE}
ex5_25 <- read.table("http://statmodel.com/usersguide/chap5/ex5.25.dat")
names(ex5_25) = paste0("y",1:12)
```

The model syntax contains a single EFA block (`efa1` for factors f1 and f2)
and single CFA block (for f3 and f4):

```{r eval=TRUE, tidy=FALSE}
model <- '
    # efa block 
    efa("efa1")*f1 + 
    efa("efa1")*f2 =~ y1 + y2 + y3 + y4 + y5 + y6

    # cfa block
    f3 =~ y7 + y8 + y9
    f4 =~ y10 + y11 + y12

    # regressions
    f3 ~ f1 + f2
    f4 ~ f3
'
```

The following command illustrates the use of various rotation arguments:

```{r eval=TRUE, tidy=FALSE, comment = ""}
fit <- sem(model = model, data = ex5_25, rotation = "geomin",
           # mimic Mplus
           information = "observed",
           rotation.args = list(rstarts = 30, row.weights = "none",
                                algorithm = "gpa", std.ov = TRUE,
                                geomin.epsilon = 0.0001))
summary(fit)
```

#### Exploratory factor analysis (EFA) ####

When there is no structural part (i.e., no regressions among the latent
variables) and there is only a single EFA block, then ESEM reduces to
exploratory factor analysis (EFA). Using the Holzinger and Swineford data,
we could specify an EFA with three factors as follows:

```{r eval=TRUE, tidy=FALSE, comment = ""}
efa.model <- '
    efa("efa")*f1 + 
    efa("efa")*f2 + 
    efa("efa")*f3 =~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9
'
fit <- cfa(efa.model, data = HolzingerSwineford1939)
summary(fit, standardized = TRUE)
```

In version 0.6-13, we added added the `efa()` function to simplify the
input, and to produce output that is more in line with traditional
EFA software in R. There is no need to create a model syntax. You 
only need to provide the data, and the number of factors. Instead
of a single number, you can also specify a range of numbers. For example:

```{r eval=TRUE, tidy=FALSE, comment = ""}
var.names <- paste("x", 1:9, sep = "")
fit <- efa(data = HolzingerSwineford1939[,var.names], nfactors = 1:3)
summary(fit)
```
