---
layout: default
title: Tutorial
published: true
submenu: tutorial
---

```{r echo=FALSE, message=FALSE}
library(lavaan)
HS.model <- ' visual =~ x1 + x2 + x3
              textual =~ x4 + x5 + x6
              speed   =~ x7 + x8 + x9 '
```
The `summary()` function gives a nice overview of a fitted model, but is for
display only. If you need the actual numbers for further processing, you may
prefer to use one of several 'extractor' functions. We have already seen the
`coef()` function which extracts the estimated parameters of a fitted model.
Other extractor functions are discussed below.

#### parameterEstimates ####

The `parameterEstimates()` function returns a data.frame containing
all the model parameters in the rows:

```{r comment=""}
fit <- cfa(HS.model, data=HolzingerSwineford1939)
parameterEstimates(fit)
```

The `lhs` (left-hand side), `op` (operator) and `rhs` (right-hand side) columns
define the parameter. The `est`, `se`, `z` and `pvalue` columns provide the
point estimate, the standard error, the z-value and the p-value for this
parameter. The last two columns are the lower and upper bounds of a 95%
confidence interval around the point estimate.

#### standardizedSolution ####

The `standardizedSolution()` function is similar to the `parameterEstimates()`
function, but only shows the standardized parameter estimates and corresponding
standard errors, z-values, p-values and confidence intervals.

#### fitted.values ####

The `fitted()` and `fitted.values()` functions return the 
model-implied (fitted) covariance matrix (and mean vector) of a fitted model:

```{r comment=""}
fit <- cfa(HS.model, data = HolzingerSwineford1939)
fitted(fit)
```

#### residuals ####

The `resid()` or `residuals()` functions return (unstandardized) residuals of a
fitted model.  This is simply the difference between the observed and implied
covariance matrix and mean vector. 

```{r comment=""}
fit <- cfa(HS.model, data = HolzingerSwineford1939)
resid(fit)
```

The `lavResiduals()` gives more extensive information about the residuals.
Per default, it will print both raw and standardized residuals, as well
as several summary statistics (including the SRMR and the unbiased SRMR).

#### vcov ####

The function `vcov()` returns the estimated covariance matrix of the parameter
estimates.

#### AIC and BIC ####

The `AIC()` and `BIC()` functions return the AIC and BIC values of a fitted
model.

#### fitMeasures ####

The `fitMeasures()` function returns all the fit measures computed by lavaan as
a named numeric vector. 

```{r comment=""}
fit <- cfa(HS.model, data=HolzingerSwineford1939)
fitMeasures(fit)
```

If you only want the value of a single fit measure,
say, the CFI, you give the name (in lower case) as the second argument:

```{r comment=""}
fit <- cfa(HS.model, data=HolzingerSwineford1939)
fitMeasures(fit, "cfi")
```

Or you can provide a vector of fit measures, as in 

```{r comment=""}
fitMeasures(fit, c("cfi","rmsea","srmr"))
```

#### lavInspect ####

If you want to peek inside a fitted lavaan object (the object that is returned
by a call to `cfa()`, `sem()`or `growth()`), you can use the `lavInspect()`
function, with a variety of options. By default, calling `lavInspect()` on a
fitted lavaan object returns a list of the model matrices that are used
internally to represent the model.  The free parameters are nonzero integers.

```{r comment=""}
fit <- cfa(HS.model, data=HolzingerSwineford1939)
lavInspect(fit)
```

To see the starting values of parameters in each model matrix, type

```{r comment=""}
lavInspect(fit, what = "start")
```
To see how lavaan internally represents a model, you can type

```{r comment=""}
lavInspect(fit, what = "list")
```

This is equivalent to the `parTable(fit)` function. The table that
is returned here is called the 'parameter table'.

For more lavInspect options, see the help page:

```{r eval=FALSE, tidy=FALSE}
?lavInspect
```
