---
layout: default
title: Tutorial
published: true
submenu: tutorial
---

```{r echo=FALSE, message=FALSE}
library(lavaan)
HS.model <- ' visual =~ x1 + x2 + x3
              textual =~ x4 + x5 + x6
              speed   =~ x7 + x8 + x9 '
```
The `summary()` function gives a nice overview of a fitted model, but is for
display only. If you need the actual numbers for further processing, you may
prefer to use one of several 'extractor' functions. We have already seen the
`coef()` function which extracts the estimated parameters of a fitted model.
Other extractor functions are discussed below.

#### parameterEstimates ####

The function extracts not only the values of the estimated parameters,
but also the standard errors, the z-values, the standardized parameter
values, as a convenient data frame. For example:

```{r comment=""}
fit <- cfa(HS.model, data=HolzingerSwineford1939)
parameterEstimates(fit)
```

#### standardizedSolution ####

The `standardizedSolution()` function is similar to the `parameterEstimates()`
function, but only shows the unstandardized and standardized parameter
estimates.

#### fitted.values ####

The `fitted()` and `fitted.values()` functions return the 
implied (fitted) covariance matrix (and mean vector) of a fitted model:

```{r comment=""}
fit <- cfa(HS.model, data=HolzingerSwineford1939)
fitted(fit)
```

#### residuals ####

The `resid()` or `residuals()` functions return (unstandardized) residuals of a
fitted model.  This is simply the difference between the observed and implied
covariance matrix and mean vector. If the estimator is maximum likelihood, it
is also possible to obtain the normalized and the standardized residuals.

```{r comment=""}
fit <- cfa(HS.model, data=HolzingerSwineford1939)
resid(fit, type="standardized")
```

#### vcov ####

The function `vcov()` returns the estimated covariance matrix of the parameter
estimates.

#### AIC and BIC ####

The `AIC()` and `BIC()` functions return the AIC and BIC values of a fitted
model.

#### fitMeasures ####

The `fitMeasures()` function returns all the fit measures computed by lavaan as
a named numeric vector. If you only want the value of a single fit measure,
say, the CFI, you give the name (in lower case) as the second argument:

```{r comment=""}
fit <- cfa(HS.model, data=HolzingerSwineford1939)
fitMeasures(fit, "cfi")
```

#### inspect ####

If you want to peek inside a fitted lavaan object (the object that is returned
by a call to `cfa()`, `sem()`or `growth()`), you can use the `inspect()`
function, with a variety of options. By default, calling `inspect()` on a
fitted lavaan object returns a list of the model matrices that are used
internally to represent the model.  The free parameters are nonzero integers.

```{r comment=""}
inspect(fit)
```

To see the starting values of parameters in each model matrix, type

```{r comment=""}
inspect(fit, what="start")
```
To see how lavaan internally represents a model, you can type

```{r comment=""}
inspect(fit, what="list")
```

For more inspect options, see the help page for the lavaan class which
you can find by typing the following:

```{r eval=FALSE, tidy=FALSE}
class?lavaan
```
