---
layout: default
title: Tutorial
published: true
submenu: tutorial
---

```{r echo=FALSE, message=FALSE}
library(lavaan)
HS.model <- ' visual =~ x1 + x2 + x3
              textual =~ x4 + x5 + x6
              speed   =~ x7 + x8 + x9 '
```

By and large, structural equation models are used to model the covariance
matrix of the observed variables in a dataset. But in some applications, it is
useful to bring in the means of the observed variables too. One way to do this
is to explicitly refer to intercepts in the lavaan syntax. This can be done by
including 'intercept formulas' in the model syntax. An intercept formula has
the following form:

    variable ~ 1

The left part of the expression contains the name of the observed or latent
variable. The right part contains the number `1`, representing the intercept.
For example, in the three-factor H&S CFA model, we can add the intercepts of
the observed variables as follows:

    # three-factor model
       visual =~ x1 + x2 + x3
      textual =~ x4 + x5 + x6
      speed   =~ x7 + x8 + x9
    # intercepts
      x1 ~ 1
      x2 ~ 1
      x3 ~ 1
      x4 ~ 1
      x5 ~ 1
      x6 ~ 1
      x7 ~ 1
      x8 ~ 1
      x9 ~ 1

However, it is more convenient to omit the intercept formulas in the model
syntax (unless you want to fix their values), and to add the argument
`meanstructure = TRUE` in the fitting function. For example, we can refit the
three-factor H&S CFA model as follows:

```{r tidy=FALSE, comment=""}
fit <- cfa(HS.model, 
           data = HolzingerSwineford1939, 
           meanstructure = TRUE)
summary(fit)
```

As you can see in the output, the model includes intercept parameters for both
the observed and latent variables. By default, the `cfa()` and `sem()` 
functions fix the latent
variable intercepts (which in this case correspond to the latent *means*) to
zero. Otherwise, the model would not be estimable. Note that the chi-square
statistic and the number of degrees of freedom is the same as in the original
model (without a mean structure). The reason is that we brought in some new
data (a mean value for each of the 9 observed variables), but we also added 9
additional parameters to the model (an intercept for each of the 9 observed
variables). The end result is an identical fit. In practice, the only reason
why a user would add intercept-formulas in the model syntax, is because some
constraints must be specified on them. For example, suppose that we wish to fix
the intercepts of the variables `x1`, `x2`, `x3` and `x4` to, say, 0.5. We
would write the model syntax as follows:

    # three-factor model
       visual =~ x1 + x2 + x3
      textual =~ x4 + x5 + x6
      speed   =~ x7 + x8 + x9
    # intercepts with fixed values
      x1 + x2 + x3 + x4 ~ 0.5*1

where we have used the left-hand side of the formula to 'repeat' the
right-hand side for each element of the left-hand side.




