---
layout: default
title: Tutorial
published: true
submenu: tutorial
---

```{r echo=FALSE, message=FALSE}
library(lavaan)
HS.model <- ' visual =~ x1 + x2 + x3
              textual =~ x4 + x5 + x6
              speed   =~ x7 + x8 + x9 '
```

Modification indices can be requested by adding the argument `modindices =
TRUE` in the `summary()` call, or by calling the function `modindices()`
directly. The `modindices()` function returns a data frame, which you can
sort or filter to extract what you want. For example, to see
only the modification indices for the factor loadings, you can use something
like this:

```{r comment="", tidy=FALSE}
fit <- cfa(HS.model, 
           data = HolzingerSwineford1939)
mi <- modindices(fit)
mi[mi$op == "=~",]
```

Modification indices are printed out for each nonfree (or nonredundant)
parameter. The modification indices are supplemented by the expected parameter
change (EPC) values (column `epc`). The last three columns contain the
standardized EPC values (`sepc.lv`: only standardizing the latent variables;
`sepc.all`: standardizing all variables; `sepc.nox`: standardizing all
but exogenous observed variables)
