---
layout: default
title: Tutorial
published: true
submenu: tutorial
---

```{r echo=FALSE, message=FALSE}
library(lavaan)
HS.model <- ' visual  =~ x1 + x2 + x3
              textual =~ x4 + x5 + x6
              speed   =~ x7 + x8 + x9 '
```

Modification indices can be requested by adding the argument `modindices =
TRUE` in the `summary()` call, or by calling the function `modindices()`
directly. By default, modification indices are printed out for each 
nonfree (or fixed-to-zero) parameter. The modification indices are 
supplemented by the expected parameter
change (EPC) values (column `epc`). The last three columns contain the
standardized EPC values (`sepc.lv`: only standardizing the latent variables;
`sepc.all`: standardizing all variables; `sepc.nox`: standardizing all
but exogenous observed variables).

A typical use of the `modindices()` function is as follows:

```{r comment = ""}
fit <- cfa(HS.model,
           data = HolzingerSwineford1939)
modindices(fit, sort = TRUE, maximum.number = 5)
```
This will print out the top 5 parameters (that can be added to the model) that 
result in the largest modification index, sorted from  high to low. 

The `modindices()` function returns a data frame, which you can
sort or filter to extract what you want. For example, to see
only the modification indices for the factor loadings, you can use something
like this:

```{r comment="", tidy=FALSE}
fit <- cfa(HS.model, 
           data = HolzingerSwineford1939)
mi <- modindices(fit)
mi[mi$op == "=~",]
```

It is important to realize that the `modindices()` function will only consider
fixed-to-zero parameters. If you have equality constraints in the
model, and you wish to examine what happens if you release all (or some) of
these equality constraints, use the `lavTestScore()` function.

