---
layout: default
title: Tutorial
published: true
submenu: tutorial
---

If the data is clustered, one way to handle the clustering is to use 
a multilevel modeling approach. In the SEM framework, this leads to
multilevel SEM. The multilevel capabilities of lavaan are still
limited, but you can fit a two-level SEM with random intercepts
(note: only when all data is continuous and complete; listwise deletion
is currently used for cases with missing values).

```{r echo=FALSE, message=FALSE}
library(lavaan)
```


#### Multilevel SEM model syntax ####

To fit a two-level SEM, you must specify a model for both levels, as 
follows:

```{r eval=TRUE, tidy=FALSE}
    model <- '
        level: 1
            fw =~ y1 + y2 + y3
            fw ~ x1 + x2 + x3
        level: 2
            fb =~ y1 + y2 + y3
            fb ~ w1 + w2
    '
```
This model syntax contains two blocks, one for level 1, and one for level 2.
Within each block, you can specify a model just like in the single-level case.
To fit this model, using a toy dataset `Demo.twolevel` that is part of
the lavaan package, you need to add the `cluster=` argument to the sem/lavaan
function call:

```{r eval=TRUE, tidy=FALSE}
    fit <- sem(model = model, data = Demo.twolevel, cluster = "cluster")
```

The output looks similar to a multigroup SEM output, but where the two
groups are now the within and the between level respectively.
```{r eval=TRUE, tidy=FALSE}
    summary(fit)
```

After fitting the model, you can inspect the intra-class correlations:
```{r eval=TRUE, tidy=FALSE}
    lavInspect(fit, "icc")
```
The see the unrestricted (h1) within and between means and covariances,
you can use

```{r eval=TRUE, tidy=FALSE}
    lavInspect(fit, "h1")
```

#### Important notes ####

- note that in `level: 1` the colon follows the `level` keyword; if you
type `level 1:`, you will get an error

- you must specify a model for each level; the following syntax is not
allowed and will produce an error:

    ```{r eval=FALSE, tidy=FALSE}
        model <- '
            level: 1
                fw =~ y1 + y2 + y3
                fw ~ x1 + x2 + x3
            level: 2
        '
    ```

- if you do not have a model in mind for level 2, you can specify a 
  saturated level by adding all variances and covariances of the 
  endogenous variables (here: y1, y2 and y3):

    ```{r eval=FALSE, tidy=FALSE}
        model <- '
            level: 1
                fw =~ y1 + y2 + y3
                fw ~ x1 + x2 + x3
            level: 2
                y1 ~~ y1 + y2 + y3
                y2 ~~ y2 + y3
                y3 ~~ y3
        '
    ```

#### Convergence issues and solutions ####

By default, the current version of lavaan (0.6) uses a quasi-Newton procedure
to maximize the loglikelihood of the data given the model (just like in the
single-level case). For most model and data combinations, this will work fine
(and fast).  However, every now and then, you may experience convergence issues.

Non-convergence is typically a sign that something is not quite right with
either your model, or your data. Typical settings are: a small number of
clusters, in combination with (almost) no variance of an endogenous variable
at the between level.

However, if you believe nothing is wrong,
you may want to try another optimization procedure. The current version
of lavaan allows for using the Expectation Maximization (EM) algorithm
as an alternative. To switch to the EM algorithm, you can use:

```{r eval=FALSE, tidy=FALSE}
    fit <- sem(model = model, data = Demo.twolevel, cluster = "cluster",
               verbose = TRUE, optim.method = "em")
```

As the EM algorithm is not accelerated yet, this may take a long time. It
is not unusual that more than 10000 iterations are needed to reach a 
solution. To control when the EM algorithm stops, you can set the stopping
criteria as follows:

```{r eval=FALSE, tidy=FALSE}
    fit <- sem(model = model, data = Demo.twolevel, cluster = "cluster",
               verbose = TRUE, optim.method = "em", em.iter.max = 20000,
               em.fx.tol = 1e-08, em.dx.tol = 1e-04)
```

The `em.fx.tol` argument is used to monitor the change in loglikelihood
between the current step and the previous step. If this change is smaller
than `em.fx.tol`, the algorithm stops. The `em.dx.tol` argument is used
to monitor the (unscaled) gradient. When a solution is reached, all 
elements of the gradient should be near zero. When the largest gradient
element is smaller than `em.dx.tol`, the algorithm stops.

A word of caution: the EM algorithm can always be forced to 'converge'
(perhaps after changing the stopping criteria), but that does not mean you 
have a model/dataset combination that deserves to converge.


  
