---
layout: default
title: Tutorial
published: true
submenu: tutorial
---

In our second example, we will use the built-in `PoliticalDemocracy` dataset.
This is a dataset that has been used by Bollen in his 1989 book on structural
equation modeling (and elsewhere). To learn more about the dataset, see its
help page and the references therein.

The figure below contains a graphical representation
of the model that we want to fit. 

```{r sem, echo=FALSE, message=FALSE, dev=c("png","pdf"), dpi=c(100,100), fig.cap="Political Democracy SEM example", fig.keep="last"}
library(lavaan) # only needed once per session
library(semPlot)
library(qgraph)
model <- '
  # measurement model
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    dem65 =~ y5 + y6 + y7 + y8
  # regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60
  # residual correlations
    y1 ~~ y5
    y2 ~~ y4 + y6
    y3 ~~ y7
    y4 ~~ y8
    y6 ~~ y8
'
fit <- sem(model, data=PoliticalDemocracy)
Graph <- semPaths(fit,
#layout=L, 
layout="tree2",
nCharNodes=0,
curve=1,
rotation=4,
#mar=c(3,3,3,3),
edge.color="black",
border.width=1.0, esize=1.0,
label.scale=FALSE,
label.cex=1.0,
residuals=FALSE,
fixedStyle=1,
freeStyle=1,
curvePivot=FALSE,
#style="LISREL",
exoVar=FALSE,
sizeMan=7,
sizeLat=10
)

L <- matrix(c(
  0.4 ,   1 ,
  0.7 ,   1 ,
  1.0 ,   1 ,
 -1.0 ,   1 ,
 -1.0 ,   0.71 ,
 -1.0 ,   0.43 ,
 -1.0 ,   0.14 ,
 -1.0 ,  -0.14 ,
 -1.0 ,  -0.43 ,
 -1.0 ,  -0.71 ,
 -1.0 ,  -1    ,
  0.7 ,   0.4 ,
  0.0 ,   0.4 ,
  0.0 ,  -0.4
), 14,2, byrow=TRUE)

qgraph(Graph, layout=L, arrowAngle=0.5)
```
The corresponding lavaan syntax for specifying this model
is as follows:

```{r eval=FALSE, tidy=FALSE}
model <- '
  # measurement model
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    dem65 =~ y5 + y6 + y7 + y8
  # regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60
  # residual correlations
    y1 ~~ y5
    y2 ~~ y4 + y6
    y3 ~~ y7
    y4 ~~ y8
    y6 ~~ y8
'
```

In this example, we use three different formula types: latent variabele
definitions (using the `=~` operator), regression formulas (using the `~`
operator), and (co)variance formulas (using the `~~` operator). The regression
formulas are similar to ordinary formulas in R. The (co)variance formulas
typically have the following form:

    variable ~~ variable

The variables can be either observed or latent variables. If the two variable
names are the same, the expression refers to the variance (or residual
variance) of that variable. If the two variable names are different, the
expression refers to the (residual) covariance among these two variables. The
lavaan package automatically makes the distinction between variances and
residual variances.

In our example, the expression `y1 ~~ y5` allows the residual variances of the
two observed variables to be correlated. This is sometimes done if it is
believed that the two variables have something in common that is not captured
by the latent variables. In this case, the two variables refer to identical
scores, but measured in two different years (1960 and 1965, respectively). Note
that the two expressions `y2 ~~ y4` and `y2 ~~ y6`, can be combined into the
expression `y2 ~~ y4 + y6`. This is just a shorthand notation.

We enter the model syntax as follows:

```{r eval=FALSE, tidy=FALSE}
model <- '
  # measurement model
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    dem65 =~ y5 + y6 + y7 + y8
  # regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60
  # residual correlations
    y1 ~~ y5
    y2 ~~ y4 + y6
    y3 ~~ y7
    y4 ~~ y8
    y6 ~~ y8
'
```

To fit the model and see the results we can type:

```{r, comment=""}
fit <- sem(model, data=PoliticalDemocracy)
summary(fit, standardized=TRUE)
```

The function `sem()` is very similar to the function `cfa()`.  In fact, the two
functions are currently almost identical, but this may change in the future. In
the `summary()` function, we omitted the `fit.measures=TRUE` argument.
Therefore, you only get the basic chi-square test statistic. The argument
`standardized=TRUE` augments the output with standardized parameter values. Two
extra columns of standardized parameter values are printed. In the first column
(labeled `Std.lv`), only the latent variables are standardized. In the second
column (labeled `Std.all`), both latent and observed variables are
standardized. The latter is often called the 'completely standardized
solution'.

The complete code to specify and fit this model is printed again below:

```{r eval=FALSE, tidy=FALSE}
library(lavaan) # only needed once per session
model <- '
  # measurement model
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    dem65 =~ y5 + y6 + y7 + y8
  # regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60
  # residual correlations
    y1 ~~ y5
    y2 ~~ y4 + y6
    y3 ~~ y7
    y4 ~~ y8
    y6 ~~ y8
'
fit <- sem(model, data=PoliticalDemocracy)
summary(fit, standardized=TRUE)
```
